Supplementary Materials for cILR: Taxonomic enrichment analysis with isometric log ratios

### 2021-12-02 cbea_revised.tex
Updates from Overleaf

However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}. First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present. For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection. For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.

Second, the total number of reads varies across samples due to limitations of the sequencing instrument \cite{quinn2018b}, where normalization approaches has to be applied to allow for comparison of feature abundances across samples. However, microbiome data sets do not follow certain assumptions that enable the usage of methods from similar fields (such as RNA-seq) \cite{quinn2019}. For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes that do not differ in expression levels across samples. As such, practitioners often rely on total sum normalization to transform count data into relative proportions constrained to sum to one \cite{weiss2017}, which based on empirical studies showed better performance in analysis tasks \cite{mcknight2019}. This normalization schema imposes spurious negative correlation between the two       
Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and is inconsistent across samples, resulting in a variable number of total read counts per sample. Many normalization methods \cite{weiss2017} have been proposed to address these issues, including cross-applying methods from the gene expression literature \cite{mcmurdie2014}. However, these methods rely on assumptions specific to the original bulk RNA-seq data sets such as the presence of housekeeping genes with consistent expression levels \cite{love2014}, which might not be not true in the context of microbiome relative abundance data \cite{quinn2019,quinn2018b}. As such, microbiome taxonomic data is strictly compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Consequently, log-ratio transformations from the compositional data literature are often utilized \cite{aitchison1999}.

Third, the data are highly zero-inflated, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment). Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015}, or applying zero-inflated models \cite{chen2018, kaul2017}. Newer methods developed recently have focused on understanding the different types of zeros in the data, providing more sophisticated heuristics around when pseudocounts can be utilized \cite{kaul2017a}.

Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}. Aggregated sets are less sparse than their constituent elements, and testing on a smaller number of variables reduces the multiple testing burden, thereby increasing power and reproducibility. Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations. There exists a diverse set of available methods developed to perform such analyses. More traditional set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest \cite{goeman2007}. Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values. The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size. Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples. Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample level and are more akin to a transformation. This strategy allows for the flexible incorporation of different statistical techniques downstream, such as prediction models, as well as for visualization purposes in ordination plots.

In microbiome research, even though no explicit enrichment analysis is performed, standard practice often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum by simple summation of abundances \cite{mclaren2019}. Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, it increases bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}. As such, there is a need for microbiome researchers to adopt more robust sample-level set enrichment methods. Unfortunately, limited work has been done to extend existing methods to be more specific to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used.

Here, we present a novel method that generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}. We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al. \cite{tian2005}, which tests the null that the value of variables within a specific set is equal to the value of measured variables not in the set. The competitive null hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables. We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a multiplicative aggregation method that addresses the downsides of the naive summation-based method presented above \cite{mclaren2019, silverman2017}. The resulting method, titled competitive isometric log-ratio (CBEA), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing. These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.

In the following sections, we provide the formulation of CBEA and discuss some statistical properties. We illustrate significance testing at the sample level using CBEA and evaluate type I error and power under different simulation scenarios and real data applications. We assess the informativeness of CBEA generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis. We compare the performance of CBEA in these respective tasks against standard microbiome taxonomic data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods. An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).

@@

\subsection*{Competitive Isometric Log-ratio (CBEA)}

@@

\subsection*{Properties of CBEA}
\subsubsection*{CBEA and the Isometric Log Ratio Transformation}
The CBEA statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}. The standard ILR is a transformation method to address the negative correlation bias inherent in compositional data by providing an isometry between the $D$-dimensional simplex $\mathbb{S}^D$ and coordinates in the $D-1$ real space $\mathbb{R}^{D-1}$ \cite{egozcue2003,washburne2017}. This is accomplished by projecting the composition onto a chosen orthonormal basis in $\mathbb{R}$, which can be defined by a sequential binary partition (SBP) of the variables (e.g. a rooted phylogenetic tree). The ILR transformed variables are the coordinates of nodes within an SBP tree of the variables. Without loss of generalizability, in a given SBP with node $i$ splitting variables between sets $R$ and $S$, we have the ILR coordinate $x^{*}_{i}$ as: 
\begin{equation}\label{ilr_standard}
    x^{*}_i = \sqrt{\frac{rs}{r+s}} \log\left(\frac{g(X_{j|j \in R})}{g(X_{j|j \in S})}\right)
\end{equation}
where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$. The ILR confer many important benefits. First, ILR coordinates exist in real space, whereby common statistical methods can be used. Second, ILR aggregated variables preserve inter-sample distances before and after aggregation \cite{egozcue2005}. Third, ILR variables are not constrained to sum to 0 as that of the centered log-ratio transformation, resulting in a covariance matrix that is not singular \cite{egozcue2003}.

@@

\subsubsection*{Statistical Properties of CBEA}
We can perform significance testing on the CBEA statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set. This is equivalent to the $Q_1$ competitive null hypothesis in the gene set testing literature \cite{tian2005} where the enrichment of a gene set is defined with respect to genes outside the set.

>>>>

However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}. The data is sparse, high-dimensional, and likely compositional \cite{gloor2017, li2019a, li2015}. Even though the these problems are challenging, a very approachable solution is to to use set-based analysis methods, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}. Aggregated variables can be less sparse, and testing on a smaller number of features can reduce the multiple-testing burden. As such, gene set testing methods have been shown to increase power and reproducibility of genomic analyses. Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more biologically informative interpretations.

There exists a diverse set of methods already developed in this field. Traditional set testing methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}. Unfortunately, these approaches are sensitive to the differential expression test as well as the chosen threshold where genes can be part of the candidate list. Aggregate score methods, which are generally more preferred \cite{irizarry2009}, instead assigns a score for each gene set based on gene-specific statistics such as z-scores or fold change. Of these approaches, methods such as GSEA \cite{subramanian2005} performs a test for each gene set at the population level, summarizing information across all samples. Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020}, generate enrichment scores at the sample level and are more akin to a transformation. In addition to being able to screen for enriched sets per sample, this strategy also allows for the flexible incorporation of different downstream analyses, such as fitting prediction models, or performing dimension reduction.

%First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present. For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016a}, allowing for simultaneous model fitting and variable selection. For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018c} to address hypothesis testing burden.

%Second, the total number of reads varies across samples due to limitations of the sequencing instrument \cite{quinn2018}.

%Third, the data are highly sparse, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment). Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015a}, or applying zero-inflated models \cite{chen2018c, kaul2017}. Newer methods developed recently have focused on understanding the different types of zeros in the data, providing more sophisticated heuristics around when pseudocounts can be utilized \cite{kaul2017a}.

In microbiome research, even though no explicit enrichment analysis is performed, researchers often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum. However, despite this interest, there is limited research done to extend existing set-based methods to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and p-value threshold. One of the primary challenges for adapting gene set analysis to the microbiome context is the compositional nature of the data. Sequencing technologies constrain the total number of reads, and samples are expected to have the same number of reads instead of DNA content \cite{quinn2019, quinn2018}. However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, suggesting the use of normalization methods to allow for proper comparison of feature abundances across samples. However, microbiome data sets do not follow certain assumptions that enable the cross-application of methods from similar fields (such as RNA-seq) \cite{quinn2019, quinn2018}. For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes with constant expression levels across samples. As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}. Empirical studies have supported this decision, showing increased performance of analysis tasks under this normalization schema \cite{mcknight2019}. Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1999} are motivated to address this issue.

Standard practice for aggregating variables in microbiome research uses element-wise summations \cite{mclaren2019}. Unfortunately, there are disadvantages to using this approach (referred to as amalgamations in the CoDA literature). First, inter-sample Aitchison distances computed on original parts are not preserved after amalgamation \cite{egozcue2005}. This means that cluster analyses might show different results depending on the level of amalgamation and differs from the those computed from original variables. Second, amalgamations do not allow for comparison between sets of different sizes within the same experimental condition since larger sets will have larger means and variances. Third, considering that each taxa has specific measurement biases \cite{mclaren2019}, an amalgamation based approach would make the bias of the amalgamated variable dependent on the relative abundance of the its constituents. In other words, if taxon $1$ has abundance $A_1$ and bias $B_1$, while taxon $2$ has abundance $A_2$ and bias $B_2$, then the bias of the aggregate variable (for example, a genera) is $(A_1B_1 + A_2B_2)/(A_1 + A_2)$ (see Appendix 1. from McLaren et al. \cite{mclaren2019}). This means that bias invariant approaches (such as analyses of ratios) would no longer be invariant when applied to amalgamated variables as bias now varies across samples. The alternative would be to multiply the proportions rather than to sum them \cite{egozcue2005}.

Here, we present a taxon-set testing method for microbiome relative abundance data that addresses the aforementioned issues. Our approach generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020}. We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al. \cite{tian2005} to formulate the enrichment of a set as the compositional balance \cite{rivera-pinto} of taxa within the set and remainder taxa using multiplication as the method of aggregating proportions \cite{egozcue2005}. This well-defined null hypothesis allows us to perform significance testing with interpretable results through estimating the empirical distribution of our statistic under the null that can also account for variance inflation due to inter-taxa correlation \cite{wu2012}.

In the following sections, we present our approach titled competitive balances for taxonomic erichment analysis (CBEA). First, we presented the step-by-step formulation of CBEA and discuss its statistical properties. Second, we detailed our evaluation strategy using both real data and parametric simulations, and the methods we're comparing against. Third, we presented results on enrichment testing using CBEA for single samples as well as at the population level. Fourth, we showed the performance of CBEA in downstream disease prediction. Finally, we discussed our results and the limitations of our method. An R package implementation of CBEA can be found on GitHub (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/CBEA}).

@@

\subsection*{Competitive balances for taxonomic enrichment analysis (CBEA)}

@@

\subsection*{Properties of CBEA}
\subsubsection*{CBEA and balances of groups of parts}
The CBEA statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}. The standard ILR is a transformation method to address the negative correlation bias inherent in compositional data by providing an isometry between the $D$-dimensional simplex $\mathbb{S}^D$ and coordinates in the $D-1$ real space $\mathbb{R}^{D-1}$ \cite{egozcue2003,washburne2017}. This is accomplished by projecting the composition onto a chosen orthonormal basis in $\mathbb{R}$, which can be defined by a sequential binary partition (SBP) of the variables (e.g. a rooted phylogenetic tree). The ILR transformed variables are the coordinates of nodes within an SBP tree of the variables. Without loss of generalizability, in a given SBP with node $i$ splitting variables between sets $R$ and $S$, we have the ILR coordinate $x^{*}_{i}$ as: 
\begin{equation}\label{ilr_standard}
    x^{*}_i = \sqrt{\frac{rs}{r+s}} \log\left(\frac{g(X_{j|j \in R})}{g(X_{j|j \in S})}\right)
\end{equation}
where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$. The ILR confer many important benefits. First, ILR coordinates exist in real space, whereby common statistical methods can be used. Second, ILR aggregated variables preserve inter-sample distances before and after aggregation \cite{egozcue2005}. Third, ILR variables are not constrained to sum to 0 as that of the centered log-ratio transformation, resulting in a covariance matrix that is not singular \cite{egozcue2003}.

@@

\subsubsection*{Estimating the null distribution}
We can perform significance testing on the CBEA statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set. This is equivalent to the $Q_1$ competitive null hypothesis in the gene set testing literature \cite{tian2005} where the enrichment of a gene set is defined with respect to genes outside the set.
